Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-01-10 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Better fig 1 refs #89
Better fig 2 refs #89
Final version, about to send
This has no major changes except for the paper split.
Last edits

Using the browser as a computing resource in volunteer computing has presented several advantages, 
but it remains a challenge to fully harness both the users’ behaviour and the browser’s capabilities.
That is why we have enhanced {\sf NodIO}, a framework used to create client-server architectures 
for distributed evolutionary algorithms, resulting in {\sf
  NodIO-W$^2$}, whose results we present in this paper.
{\sf NodIO-W$^2$} leverages the HTML5 Web Workers specification to execute in the background several clients per
browser, so that the the number of generations per user could be improved. The enhancements on the NodIO architecture,
along with some initial results on the expected performance and scaling behavior when the number of users is increased,
are introduced. We also use the algorithm data to find out what kind of patterns the user behavior follows by examining 
the resources they devote to our experiment, and try to fit them in some statistical distribution. In general, 
our experiments prove, first, that {\sf NodIO} is a flexible and low-overhead platform to perform distributed computation
experiments that include volunteers, and then that user behavior
follows some patterns, and, thus, are amenable to a model and thus predictable.
% Volunteer computing is extensively used in science and industry, but
% it still has a long way to go in order to become a mainstream
% technology in the same way the cloud technologies in general are. %I have changed "have" to "are"
% The lack of general purpose and open
% source tools with built-in analytics could be one of the reasons of
% why this is so, as well as the lack of proofs of concept.
% % (Paloma) The following sentence is way too long, and the "which" after "Javascript" is confusing, so one does not know if the which refers to the framework or Javascript by reading the whole sentence.
% % I propose: "That is why in this paper we present NodIO, a framework used to create client-server architectures for distributed evolutionary algorithms, entirely written in JavaScript. Our framework needs minimal server support to work and is able to support persistent, asynchronous, distributed evolutionary algorithms running in the browser, as well as any other client consuming its REST application programming interface."
% That is why
% in this paper we present {\sf NodIO}, a framework used to create
% client-server architectures for distributed evolutionary algorithms,
% entirely written in JavaScript, which needs minimal server support to
% work and that is able to support persistent, asynchronous, distributed
% evolutionary algorithms running in the browser, as well as any other
% client consuming its REST application programming interface. An
% enhanced implementation, which we have called {\sf NodIO-W$^2$}, is
% also presented; this version leverages the HTML5 Web Workers
% specification to execute several clients per browser and do it in the
% background, so that the the number of generations per user could be
% improved. The architecture, along with some initial results on the
% expected performance and scaling behavior when the number of users is
% increased is introduced. % (Paloma) I've changed this "presented" for a synonym
% We also use the algorithm data to find out
% what kind of patterns the user behavior follows by examining the
% resources they devote to our experiment, and try to fit them in % (Paloma) Changed. Phrasal verb was wrong.
%  some
% statistical distribution to create the bases of a model of user
% behavior in volunteer computing experiments of this kind that can be
% used by researchers to predict their performance. %(Paloma) Changed to "their" because of "experiments"
% In general, our
% experiments prove, first, that {\sf NodIO} is a flexible and
% low-overhead platform to perform distributed computation experiments
% that include volunteers, and then that user behavior follows some
% patterns, and, thus, are amenable to a model.
\end{abstract}

@@

%Mario: Is NodIO presented here after the cut? 
%TO DO: Add reference
There is, then, space for frameworks that do not make any of these
assumptions, such as {\sf NodIO} presented in \cite{}. {\sf NodIO} is basically 
%assumptions, such as the one presented in this paper, which will be
% named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can be embedded in any web-enabled device, including many over which 
developers have no control.

@@

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or by farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of gray indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.png 
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.png 
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

>>>>

Using the browser as a computing resource in volunteer computing has presented several advantages, 
but it remains a challenge to fully harness both the users' behaviour and the browser's capabilities.
That is why we have enhanced {\sf NodIO}, a framework used to create client-server architectures 
for distributed evolutionary algorithms, resulting in {\sf
  NodIO-W$^2$}, whose results we present in this paper.
{\sf NodIO-W$^2$} leverages the HTML5 Web Workers specification to execute in the background several clients per
browser, so that the the number of generations per user could be improved. The enhancements on the NodIO architecture,
along with some initial results on the expected performance and scaling behavior when the number of users is increased,
are introduced. We also use the experimental data to find out what kind of patterns the user behavior follows by examining 
the resources they devote to our experiment, and try to fit them in some statistical distribution. In general, 
our experiments prove, first, that {\sf NodIO} is a flexible and low-overhead platform to perform distributed computation
experiments that include volunteers, and then that user behavior
follows some patterns, and, thus, are amenable to a model and thus predictable.
% Volunteer computing is extensively used in science and industry, but
% it still has a long way to go in order to become a mainstream
% technology in the same way the cloud technologies in general are. %I have changed "have" to "are"
% The lack of general purpose and open
% source tools with built-in analytics could be one of the reasons of
% why this is so, as well as the lack of proofs of concept.
% % (Paloma) The following sentence is way too long, and the "which" after "Javascript" is confusing, so one does not know if the which refers to the framework or Javascript by reading the whole sentence.
% % I propose: "That is why in this paper we present NodIO, a framework used to create client-server architectures for distributed evolutionary algorithms, entirely written in JavaScript. Our framework needs minimal server support to work and is able to support persistent, asynchronous, distributed evolutionary algorithms running in the browser, as well as any other client consuming its REST application programming interface."
% That is why
% in this paper we present {\sf NodIO}, a framework used to create
% client-server architectures for distributed evolutionary algorithms,
% entirely written in JavaScript, which needs minimal server support to
% work and that is able to support persistent, asynchronous, distributed
% evolutionary algorithms running in the browser, as well as any other
% client consuming its REST application programming interface. An
% enhanced implementation, which we have called {\sf NodIO-W$^2$}, is
% also presented; this version leverages the HTML5 Web Workers
% specification to execute several clients per browser and do it in the
% background, so that the the number of generations per user could be
% improved. The architecture, along with some initial results on the
% expected performance and scaling behavior when the number of users is
% increased is introduced. % (Paloma) I've changed this "presented" for a synonym
% We also use the algorithm data to find out
% what kind of patterns the user behavior follows by examining the
% resources they devote to our experiment, and try to fit them in % (Paloma) Changed. Phrasal verb was wrong.
%  some
% statistical distribution to create the bases of a model of user
% behavior in volunteer computing experiments of this kind that can be
% used by researchers to predict their performance. %(Paloma) Changed to "their" because of "experiments"
% In general, our
% experiments prove, first, that {\sf NodIO} is a flexible and
% low-overhead platform to perform distributed computation experiments
% that include volunteers, and then that user behavior follows some
% patterns, and, thus, are amenable to a model.
\end{abstract}

@@

%Mario: Is NodIO presented here after the cut? 
%TO DO: Add reference
There is, then, space for frameworks that do not make any of these
assumptions, such as {\sf NodIO} presented in \cite{2016arXiv160101607M}. {\sf NodIO} is basically 
%assumptions, such as the one presented in this paper, which will be
% named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can be embedded in any web-enabled device, including many over which 
developers have no control.

@@

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or by farming out evaluations to all
available nodes. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of gray indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande
